---
title: "Turning images into ridgeline plots"
author: "Atrebas"
date: "2019-05-18"
summary: I was wondering what images turned into ridgeline plots would look like. This post illustrates some experiments performed with ggplot2 and ggridges.     
output:
  blogdown::html_page
categories: ["R"]
tags: ["R", "ggplot2"]
---

<style>
  .hljs{
    background: #a7a7a71a;
    font-size: 90%;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


What if we turn images into ridgeline plots?    

# Ridgeline plots


> "Ridgeline plots are partially overlapping line plots that create the impression of a mountain range. They can be quite useful for visualizing changes in distributions over time or space." - Claus Wilke.

They have been quite popular recently. Some references include:

- The work of [James Cheshire](http://spatial.ly/2014/08/population-lines/) to represent population density    
- Alex Whan's [post](http://alexwhan.com/2016-03-24-joy-division-plot), reproducing the Joy Division's album cover    
- The [ggridges](https://github.com/clauswilke/ggridges) package from Claus Wilke    
- In Python, the recent [ridge_map](https://github.com/ColCarroll/ridge_map) library by Colin Carroll finally convinced me to perform some experiments     
     


# Import images

First, let's load the required packages. Then, we create a function to import an image from a url
and store the grayscale pixel values as a matrix.    


```{r, message = FALSE}

library(data.table)
library(ggplot2)
library(ggridges)
library(jpeg)

img_to_matrix <- function(imgurl) {
  tmp <- tempfile()
  download.file(imgurl, tmp, mode = "wb")
  img <- readJPEG(tmp)
  file.remove(tmp)
  img <- t(apply(img, 2, rev)) # rotate
  print(paste("Image loaded.", nrow(img), "x", ncol(img), "pixels."))
  img
}

```

We'll use two images throughout this post.

```{r}
url1 <- "http://upload.wikimedia.org/wikipedia/en/8/86/Einstein_tongue.jpg"
url2 <- "http://upload.wikimedia.org/wikipedia/commons/thumb/3/3b/Monroe_1953_publicity.jpg/391px-Monroe_1953_publicity.jpg"

img1 <- img_to_matrix(url1)
img2 <- img_to_matrix(url2)

par(mfrow = c(1, 2))
image(img1, col = gray.colors(50))
image(img2, col = gray.colors(50))
```


## Convert image to a data.frame

Next, the `matrix_to_dt` function will convert an image matrix into a data.frame. The data.table package is used here for convenience.    
This function is simply used to melt the matrix. The coordinates on the y axis that are used below (`y2`) correspond to the y value + the pixel intensity (`z`).     

This function has four parameters:     

- `img`: the matrix object     
- `ratio`: integer value, used to reduce the size of the matrix     
- `height`: numeric value, scaling applied to pixel value, higher will make peaks taller (and create overlap)    
- `y_as_factor`: boolean, will make things easier with ggplot2 `geom_ribbon()`    


```{r}
matrix_to_dt <- function(img, ratio = NULL, height = 3, y_as_factor = FALSE) {
  if (!is.null(ratio)) {
    img <- img[seq(1, nrow(img), by = ratio), 
               seq(1, ncol(img), by = ratio)]
  }
  
  imgdt <- data.table(x = rep(1:nrow(img), ncol(img)),
                      y = rep(1:ncol(img), each = nrow(img)),
                      z = as.numeric(img))
  
  imgdt[, y2 := y + z * height]
  
  if (y_as_factor) {
    imgdt[, y := factor(y, levels = max(y):1)]
    setorder(imgdt, -y)
  }
  
  imgdt[]
}
```


## Basic plots

Starting with a simple ggplot2 plot, using `geom_path()`.

```{r fig.height = 8.7}
imgdt <- matrix_to_dt(img1, ratio = 2L)

ggplot(data = imgdt) +
  geom_path(aes(x     = x,
                y     = y2,
                group = y),
            size = 0.15) + 
  theme_void()
```

Here is another try increasing the height.    

```{r fig.height = 8.7}
imgdt <- matrix_to_dt(img1, height = 5L, ratio = 2L)

ggplot(data = imgdt) +
  geom_path(aes(x     = x,
                y     = y2,
                group = y),
            size = 0.2) + 
  theme_void()
```

```{r fig.height = 8.6}
imgdt <- matrix_to_dt(img2, height = 5L, ratio = 4L)

ggplot(data = imgdt) +
  geom_path(aes(x     = x,
                y     = y2,
                group = y),
            size = 0.2) + 
  theme_void()
```


We can also used vertical lines. In fact, I think they are more appropriate for these pictures.    


```{r fig.height = 8.7}
imgdt <- matrix_to_dt(img1, height = 5L, ratio = 2L)

ggplot(data = imgdt) +
  geom_path(aes(x     = x + z * 5L,
                y     = y,
                group = x),
            size   = 0.15) + 
  theme_void()
```


```{r fig.height = 8.6}
imgdt <- matrix_to_dt(img2, height = 5L, ratio = 3L)

ggplot(data = imgdt) +
  geom_path(aes(x      = x + z * 4L,
                y      = y,
                group  = x,
                colour = x),
            size = 0.2) +
  scale_colour_continuous(type = "viridis", guide = FALSE) +
  theme_void()
```


## Ribbon plots

To prevent the lines from overlapping, we can use `geom_ribbon`. The y values are converted into factors to keep them in the right order.   
    
    
```{r fig.height = 8.7}
imgdt <- matrix_to_dt(img1, height = 5L, y_as_factor = TRUE, ratio = 2L)

ggplot(data = imgdt) +
  geom_ribbon(aes(x     = x,
                  ymax  = y2,
                  ymin  = 0,
                  group = y),
              size   = 0.15,
              colour = "white",
              fill   = "black") + 
  theme_void()
```

```{r fig.height = 8.6}
imgdt <- matrix_to_dt(img2, height = 7L, y_as_factor = TRUE, ratio = 4L)

ggplot(data = imgdt) +
  geom_ribbon(aes(x     = x,
                  ymax  = y2,
                  ymin  = 0,
                  group = y),
              size   = 0.15,
              colour = "white",
              fill   = "black") + 
  theme_void()
```

## ggridges

The plots above are a bit sharp and lack the smooth aspect of ridgeline plots.    
A first solution could be to use a smoothing spline. As an alternative, I gave a try to ggridges. The trick here is to transform the data by repeating the x values proportionally to the pixel intensity.    


```{r fig.height = 8.7}
imgdt <- matrix_to_dt(img1, height = 7L, y_as_factor = TRUE, ratio = 2L)

imgdt2 <- imgdt[, .(x = rep(x, round(z * 100))), by = y]
imgdt2[, y := factor(y, levels = rev(levels(y)))]

ggplot(imgdt2,
       aes(x = x,
           y = y)) +
  stat_density_ridges(geom      = "density_ridges", 
                      bandwidth = 0.8,
                      colour    = "white",
                      fill      = "black",
                      scale     = 7,
                      size      = 0.15) +
  theme_ridges() +
  theme_void() +
  theme(panel.background = element_rect(fill = 'black')) 
```


```{r fig.height = 8.6}
imgdt <- matrix_to_dt(img2, height = 7L, y_as_factor = TRUE, ratio = 4L)

imgdt2 <- imgdt[, .(x = rep(x, round(z * 100))), by = y]
imgdt2[, y := factor(y, levels = rev(levels(y)))]

ggplot(imgdt2,
       aes(x = x,
           y = y)) +
  stat_density_ridges(geom      = "density_ridges", 
                      bandwidth = 0.8,
                      colour    = "white",
                      fill      = "black",
                      scale     = 5,
                      size      = 0.2,
                      rel_min_height = 0.15) +
  theme_ridges() +
  theme_void() +
  theme(panel.background = element_rect(fill = 'black'))
```

```{r fig.height = 8.6}
ggplot(imgdt2,
       aes(x = x,
           y = y)) +
  stat_density_ridges(geom      = "density_ridges", 
                      bandwidth = 0.8,
                      alpha     = 1,
                      size      = 0.01,
                      fill      = "white",
                      color     = NA,
                      rel_min_height = 0.15) +
  theme_ridges() +
  theme_void() + 
  theme(panel.background = element_rect(fill = 'black'))
```


<br>

## Conclusion

Zooming in and out the high-resolution plots rendered as a pdf file is mesmerizing and it is quite impressive to see the amount of details captured by these few intertwining lines.    

Here is a final example.     

```{r, message = FALSE, warning = FALSE}
jpeg("tmp.jpg")
plot(0, 0, type = "n", axes = FALSE, xlab = "", ylab = "")
text(0, 0, "ggplot2\nrocks!", cex = 9)
dev.off()

img   <- readJPEG("tmp.jpg")
img   <- 1 - t(apply(img[, , 1], 2, rev))
imgdt <- matrix_to_dt(img, height = 4L, y_as_factor = TRUE, ratio = 3L)
file.remove("tmp.jpg")

ggplot(imgdt) +
  geom_ribbon(aes(x      = x,
                  ymax   = y2,
                  ymin   = 0,
                  group  = y,
                  colour = y),
              size = 0.6,
              fill = "white") + 
  theme_void() +
  scale_colour_discrete(guide = FALSE)
```


<br>


```{r}
sessionInfo()
```


